On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-08 paper.tex

\section{Introduction}

\input{model}
\input{linear}
\input{result}
%\input{linear_adia}
%\input{numerical}
%\input{application} 
\input{summary}

>>>>

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas drag introduces a 
wide range of new phenomena that are absent in pure gas
disks. Important examples include the streaming instability
\citep{youdin05a,youdin07b,johansen07} and secular gravitational
instabilities \citep{ward00,youdin11,takahashi14} that lead to
planetesimal formation. More recently, new dusty instabilities have
also been uncovered \citep{loren15,loren16,lamb16}

%Understanding dusty gas dynamics is
%therefore essential to any planet formation theory. 
%phenomena that depends on gas drag 
%planet formation in dusty disks 
%dust component is building block of planets
%understanding dusty gas dynamics 
%small amount of dust can have significant consequences

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics with explicit
Lagragian dust particles \citep{nelson10,bai10,johansen11,yang14,zhu14,gibbons15,simon16,baruteau16}. In this
approach, the equations of motion for each particle is solved
including dust-gas drag, the strength of which is measured 
by a stopping time $\tstop$ --- the characteristic timescale for the
relative velocity between gas and dust to decay. Thus small $\tstop$
corresponds to tightly-coupled particles.

%refs

Another approach is to model the dust population as a continous, pressureless
fluid
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The equations
hydrodynamics are solved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$,
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of, and adapt, already-developed methods/codes for
simulating pure gas dynamics.

%refs

It is possible to re-formulate the two-fluid equations into that for a
single fluid of for the total density $\rho$ and center-of-mass
velocity $\bm{v}$ of the dusty gas mixture
\citep{laibe14}. The standard hydrodynamic equations for
$\rho$ and $\bm{v}$ are supplemented by evolutionary equations for
the dust-to-gas ratio $\rhod/\rhog$ and the dust-gas velocity
difference $\Delta\bm{v}$. The latter also contributes to
additional source terms in the momentum and energy equations.  
The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can make the two-fluid or Lagragian particles approach
numerically challanging; but is straight forward to model in the
one-fluid framework \citep{price15}.

The single fluid formalism also has the important advantage that it is
amenable to standard analytical methods
\citep{youdin05a,jacquet11}. This is the necessary  approach to gain physical insight.  
In this work, we are motivated to develop a general analytical
framework to  study the stability of dusty protoplanetary disks by
drawing on the physical and mathematical parallel between dusty gas and standard
hydrodynamics. This is possible in certain limits of astrophysical
interest.

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a,
  jacquet11,laibe14}. We also consider irradiated protoplanetary disks 
in which the temperature is set externally \citep{chiang97,stam08}. We
can then adopt a locally isothermal equation of state $P = c_s^2\rhog$
where $c_s$ is a prescribed sound-speed profile.  This physically
represents short cooling times \citep{lin15}.

Under these approximations, we show that the one-fluid equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas; and this property is analogous to the advection
of entropy following an adiabatic fluid. We will formally relate the
entropy of isothermal dusty gas to its dust content. Finite 
dust-gas drag can be re-interpreted as heating/cooling: the relative
motion between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

%suitable for 
%analytic studies 
%more importantly, can understood

By re-casting the dusty problem into identical form to standard
hydrodynamics, we can effortlessly generalize previous stability
results for pure gas disks to the dusty case. We demonstrate this for
perfectly coupled dust in strictly and locally isothermal disks. We
show that sharp dust edges may destabilize protoplanetary disks.  
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}. We provide 
thermodynamical interpretations of our dusty stability calculations.

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in
fact the usual hydrodynamic equations, including a precise 
definition of the entropy isothermal dusty gas. We describe approximate 
equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty problem in various
limits. We use these results to constraint the sharpness of dust edges
in isothermal disks. In \S\ref{linear_problem} we formulate an
explicit stability calcultion to study the dusty VSI, for which we
present numerical results \S\ref{results}. We summarize and discuss
future applications of the dusty/adiabatic gas equivalence in
\S\ref{summary}.

\input{model}
\input{linear}
\input{result}
\input{summary}

### 2016-08-08 summary.tex
rough intro

\section{Summary and discussion}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. 
In the limit of perfect dust-gas coupling, the dust content of a fluid 
parcel $\rhod/\rhog$ is conserved. This is analogous to entropy being 
conserved following a adiabatic, pure gas. %s = s(\rhod/\rhog)
We explicitly show that for 
a locally isothermal, dusty gas that the evolutionary equation of 
$\rhod/\rhog$ may be recast into an effective energy equation identical
in form to that in standard adiabatic hydrodynamics. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
This allows us to define the buoyancy frequency of an isothermal 
dusty gas.

@@

%\subsection{Implication for numerical simulations}

\subsection{Thermodynamic interpretation of dusty instabilities}

%\subsection{Caveats and future directions}
%gi

>>>>

\section{Summary and discussion}\label{summary}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. 
In the limit of perfect dust-gas coupling, the dust content of a fluid 
parcel $\rhod/\rhog$ is conserved. This is analogous to entropy being 
conserved following a adiabatic, pure gas. %s = s(\rhod/\rhog)
We explicitly show that for 
a locally isothermal, dusty gas that the evolutionary equation of 
$\rhod/\rhog$ may be recast into an effective energy equation identical
in form to that in standard adiabatic hydrodynamics. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
This allows us to define the buoyancy frequency of an isothermal 
dusty gas.

@@

\subsection{Thermodynamic view of dusty instabilities}

%streaming instability 
Finite dust-gas drag can lead to the streaming instability (SI) in the
presence of a background radial pressure gradient \citep{youdin05a}.    
SI can also be recovered from the one-fluid framework
\citep{laibe14}. \cite{jacquet11} explains the essence of SI as dust
accumulation at pressure maxima (from terminal velocity
approximation), which back-reacts on the gas flow in 
such a way to strengthen said maxima.

% Now consider our energy formulation of the
% dusty problem, Eq. \ref{eff_energy}. Assume constant $c_s$
% and that the dominant contribution to the energy source
% $\mathcal{C}$ is associated with $\nabla^2P$ \citep[which is
% responsible for SI, ][]{jacquet11}. Then 
% \begin{align*}
% \frac{DP}{Dt} \simeq \frac{P}{\rho}\frac{D\rho}{D t} + c_s^2\tepsilon \tstop
%   \nabla^2 P.
% \end{align*} 
% At pressure maxima, dust-drag acts as a cooling term. However, if
% dust accumulation into this maximum leads to a sufficiently large
% compression, there may be a net heating, enhancing the pressure
% maximum, and a positive feedback established.

% This picture relies on unstable correlations between pressure
% perturbations and fluid compression, as pointed out by
% \cite{jacquet11}. Indeed, we see from Eq. \ref{thermal_instabilities}
% that instabilities due to drag are attributed to  
% \begin{align*}
%   \left(\nabla\cdot\dd\bm{v}^*\right) \delta \mathcal{C}.
% \end{align*} 
% Thus instabilities may arise if $\nabla\cdot\delta \bm{v}$ is suitably
% correlated to the perturbed drag force $\delta \mathcal{C}$ (which for
% SI is due to pressure perturbations). This is analogous  
% to thermal instabilities in gas dynamics where $\mathcal{C}$
% represents non-adiabatic heating/cooling and may lead to

% \citep{latter06},

%loren's instability
